\large How resource abundance and stochasticity affect animals' spatial needs

### 2023-08-06 writing/manuscript.Rmd
updated based on Lauren's comments; fixed fig pos

In this paper, we refer to a location's average amount of resources as **resource abundance**, while use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use. For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies and life histories being selected for. Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

@@

When $\e(R)$ is high, we expect animals' spatial requirements be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan. For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones. For example, the abundance of berries is predictably scarce outside the growing season, while it is higher and unpredictable during spring and summer, since a bear may explore a location only after competitors ate all the berries or during a year of unusually low productivity. In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].

@@

```{r stoch-timescale, eval = FALSE, fig.cap="Conceptualizations of perceptions of $\\e(R)$ (a.) and $\\var(R)$ (b.) over time for animals with no plasticity in their movement (blue), seasonal memory but no plasticity (orange), or high plasticity but no memory (green). The true trends in $\\e(R)$ and $\\var(R)$ are indicated in black in panels (a.) and (b.), respictively. Estimates in the white area indicate the animal's estimated $\\mu(t)$ and $\\sigma^2(t)$ based on past experience, while estimates to the grey shaded region indicate the animals' predictions of $\\mu(t)$ and $\\sigma^2(t)$ in the future. The four facets show how each behavior performs in environments where $\\e(R)$ is either predictable (top row) or stochastic (bottom row), and $\\var(R)$ is either predictable (left column) or stochastic (right column). Animals that do not adapt to changes in $R$ (blue) are unable to predict $\\e(R)$ accurately, and they have high estimates of $\\var(R)$ when $\\mu(t)$ is far from the long-term, temporally static average. Memory of seasonal trends in $\\mu(t)$ and $\\var(R)$ (orange) allows animals to perform well in predictable environments, but their lack of adaptability causes them to fair poorly in stochastic habitats, where they perceive $\\var(R)$ to be high when $\\e(R)$ is far from the expected $\\mu(t)$. Finally, the ability to adapt rapidly to change (green) allows animals to track changes in $\\e(R)$ and $\\var(R)$ closely, but the lack of memory prevents accurate predictions in predictable environments. Since such animals tend to perceive changes in $R$ as changes in $\\mu(t)$, they tend to have lower estimates of $\\sigma^2(t)$ than other animals."}
# set eval to false: include in a separate paper?
knitr::include_graphics('../figures/stochasticity-timescale.png')
```

@@

\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e. its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary. `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs. Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).

@@

In contrast, individuals with short lifespans are less likely to experience stochastic events and develop memory about them. However, the short generation time and high reproductive rate of r-selected species [as opposed to k-selected species, see @pianka_r-_1970; @brown_toward_2004] may promote other traits that increase survival following extreme events. Examples include higher population size, dispersal, higher genetic diversity, and complex group dynamics [e.g., @cuadrado_assessing_2021; @stefanescu_timing_2021]. Events that occur over large spatiotemporal scales can thus promote slow, large-scale changes through the adaptation, natural selection, and evolution of the species [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which is more likely to adapt if starting populations are large with high genetic diversity and short generations, as in the case of r-selected species [but see @leung_smaller_2022]. Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than lower temperatures at night than during the day and some weather stochasticity. Still, a population may respond to spatiotemporal cycles in $\e(R)$ and $\var(R)$ by adapting the group's reproduction timing and migration [e.g., painted lady butterflies: @cuadrado_assessing_2021; @stefanescu_timing_2021].

>>>>

In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use. For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies and life histories being selected for. Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

@@

When $\e(R)$ is high, we expect animals' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005]. However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019]. It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020]. Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan. For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].

@@

\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021]. However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones. For example, since berries are scarce outside the growing season, the variance in berries will also be low. But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location only after competitors ate all the berries. In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].

@@

@@

\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e. its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary. `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs. Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).

@@

In contrast, individuals with short lifespans are less likely to experience stochastic events and develop memory about them. However, the short generation time and high reproductive rate of r-selected species [as opposed to k-selected species, see @pianka_r-_1970; @brown_toward_2004] may promote other traits that increase survival following extreme events. Examples include higher population size, dispersal, higher genetic diversity, and complex group dynamics [e.g., @cuadrado_assessing_2021; @stefanescu_timing_2021]. Events that occur over large spatiotemporal scales can thus promote slow, large-scale changes through the adaptation, natural selection, and evolution of the species [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which is more likely to adapt if starting populations are large with high genetic diversity and short generations, as in the case of r-selected species [but see @leung_smaller_2022]. Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than daily cycles and some weather stochasticity. Still, a population may respond to spatiotemporal cycles in $\e(R)$ and $\var(R)$ by adapting the group's reproduction timing and migration [e.g., painted lady butterflies: @cuadrado_assessing_2021; @stefanescu_timing_2021].
